Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-06-30 agujournaltemplate.tex
Ago revision completed all comments/corrections

\begin{keypoints}
\item The CARIACO phytoplankton community time series reveals two clusters, separated by a shift from high- to low-upwelling conditions after 2004 and a return to high-upwelling conditions following 2013.
\index No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low-upwelling conditions.
\item Gradient forest analysis identifies the AMO climate index as the strongest predictor of community change, highlighting the influence of large-scale climatic oscillations on this tropical coastal ecosystem.
\end{keypoints}

@@

\begin{abstract}
Phytoplankton communities play a crucial role in oceans' biogeochemical cycles. Climate drives changes in the physical environment that phytoplankton communities experience, thus impacting the productivity and diversity of marine ecosystems. Long-term time series can capture changes in climate and ecosystems, allowing us to untangle the effects on local communities. 
This study aims to describe and investigate the changes in phytoplankton community in the Cariaco Basin, Venezuela, between 1995 and 2017. Water samples were taken during monthly cruises to determine microscopy cell counts and biogeochemical properties, which we integrated with data from the AMO and MEI v.2 climate indices and ERA5 climate model reanalysis data. 
Using clustering algorithms, we investigate the structure of the community data, revealing two clusters that describe a shift from high- to low-upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014.
However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009. To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related variables (SST and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors. Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of ecosystems to recover from the effects of climate change.

@@

    A tropical coastal location that has been studied intensively in the last decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters have been consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}. 
    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world oceans \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression comprising two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated. This underwater topography in combination with a high surface productivity and aerobic degradation of sinking organic matter, creates permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for understanding the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.

@@

    The monthly AMO index data (10-year low-pass), which represents an area-averaged, detrended low-pass filtered North Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.

@@

\subsection{Data analysis}    
    To provide an overview of the data available and the general trends shown by the data, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$). Cruises from 1995, 2016, and 2017 were excluded because insufficient coverage in these years prevented us from producing reliable yearly mean estimates.

@@

To check whether the annual aggregation of community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (Figure \ref{fig:clustering}$c$). The NMDS clustering is overlaid with environmental data and climate indices in the attempt of unveiling some of the underlying trends. The horizontal axis separates the two clusters, with cluster 2 located over positive anomalies in AMO and precipitation and cluster 1 located over positive anomalies in evaporation, salinity, and MEI v.2. The second NMDS axis appears to follow the trends in upwelling intensity, as indicated by the vectors pointing in opposite directions for nutrient concentrations, temperature, and wind speed. Overall, the NMDS clustering confirms the pattern shown by clustering on annually aggregated data.

@@

\subsection{Community reorganization and biomass shifts}
% P1: Introductory paragraph highlighting the overall dynamics/shifts
This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations impacted the wind-driven upwelling regime, producing confounding effects on the composition and biodiversity of the phytoplankton community.
These changes are well visible in the yearly mean anomalies (Figure \ref{fig:zscore}) and the density distribution of monthly measurements (Figure \ref{fig:clustering}). After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident to a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies up to 2014. 
By considering the time series data until the end of 2017, the temporally complete dataset, we find that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}. In 2014, physical conditions in the Cariaco Basin returned to a regime of increased wind speed, increased upwelling, and elevated nutrient concentrations (Figure \ref{fig:zscore}). Concurrently, we observe an increase in chlorophyll \textit{a} (Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (Figure \ref{fig:clustering} $a$ and $c$).

% P2: Discuss community reorganisation, biomass shifts vs HPLC data
The annual aggregates of community data showed two regimes, one covering the periods 1996 to 2004 and 2013-2017 and the other covering the period between 2004 and 2013 (see Figure \ref{fig:clustering}).  
When comparing the two regimes with respect to phytoplankton community data, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a}, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}). The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community might have transitioned to smaller cells, which sizes were probably below the detection limit of optical microscopes \cite{muller-karger_scientific_2019}. This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}. Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but shifts towards smaller cells are particularly prevalent when temperatures increase and nutrient supply decreases \cite{mousing_global_2014}, as we observed in the Cariaco Basin.

% Disucssion of HPLC vs Microscopy!
In addition to our focus on the most temporally complete dataset, another important difference between our work and previous analyses of the Cariaco time series like, for example, the study of \citeauthor{pinckney_phytoplankton_2015}, is that, for unveiling changes in the community composition of phytoplankton, we concentrated on  microscopy counts, whereas  \citeauthor{pinckney_phytoplankton_2015} considered HPLC pigment data. The patterns we found in the microscopy data are less clear in the HPLC pigment data, which instead show an increase in chlorophyll \textit{a} between the periods 1996-2000 and 2006-2010, particularly at depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}. Cell counts of all functional groups decrease, while HPLC data in  \citeauthor{pinckney_phytoplankton_2015} indicate an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{meters} . It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006. Additionally, \citeauthor{pinckney_phytoplankton_2015} could not use a unique, shared method for quantifying chlorophyll \textit{a} across different laboratories. 
Here, we focus on consistent fluorometric measurements. Consistently with our results, independent spectrophotometric measurements of chlorophyll \textit{a} concentrations at \qty{20}{meters} depth, off the coast of Margarita Island, show a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}. Changes in the composition of the phytoplankton community can affect chlorophyll \textit{a} concentrations, as most measurement methods do not capture only chlorophyll \textit{a} but also other related forms of chlorophyll, which can vary depending on the composition of the phytoplankton community \cite{welschmeyer_fluorometric_1994}.

% P3: Shift in Diversity does not match shift in biomass/cell counts, why?
The shift we find in yearly mean genus richness, however, does not correspond with abundances in cell counts. Instead, we observe a marked reduction from 1998 to 2010, after which richness recovers four years prior to 2014, when the community returns, seemingly, to previous levels, those characteristics of regime 1. The same trend is evident, albeit to a lesser degree, in the other diversity metrics. Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and exhibit a reduced yearly mean between 2000 and 2006, during which cell counts recorded an increase in nanoflagellates and a pronounced peak in cyanobacteria.
 We observe a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly an increase a few years prior to the return in abundance to previous levels. Microscopy cell counts do not show any statistical difference in diversity between the two  regimes, either by Shannon index or by genus richness (Figures \ref{fig:zscore} and \ref{fig:divts}).
We observe a significant difference in the Pielou's index between the two regime. The HPLC pigment trend unveiled by and Our analysis on microscopy measurements suggesting that the phytoplankton community becomes more evenly distributed during low-upwelling conditions of regime 2 are consistent with the analysis of \citeauthor{pinckney_phytoplankton_2015} based on HPLC pigment data. This increase in evenness, quantified by the Pielou index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community. The resulting decrease in compositional turnover is also observed in a decline in interannual species turnover during regime 2 (see Figure \ref{fig:clustering}$b$). As the productivity and biomass of phytoplankton communities are driven primarily by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.

@@

% P5: Climate impacts
Although regional effects may vary, a negative phase in the AMO is generally associated with periods of colder temperatures and a southward shift of the ITCZ \cite{knight_climate_2006, colna2017latitudinal}. In the Cariaco Basin, a southward shift of the ITCZ correlates with increased wind-driven upwelling \cite{taylor_ecosystem_2012}. The threshold responses between functional groups, which relate to shifts in cell count abundances, show contrasting patterns. Diatoms, dinoflagellates, and haptophytes exhibit strong response thresholds for negative AMO anomalies, which could point to influences of stronger upwelling conditions. The abundance of haptophytes shows, additionally, a threshold effect at positive AMO anomalies, which may be related to the pattern observed by \citeauthor{pinckney_phytoplankton_2015} using HPLC data of haptophytes predominantly exhibiting blooms above the mixed layer and at a lower frequency than other functional groups.

The El Niño Southern Oscillation (ENSO) exhibits a strong influence on the Caribbean Sea. During ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months. In contrast, we found a relatively strong predictive power for changes in the phytoplankton community for the MEI v.2 index with a 4-month time lag. We also tested a 12-month lag and found no significant improvement (Table \ref{tab:sup:GFlagtests} in Appendix).

The threshold responses indicate a shift in the abundances of dinoflagellates and haptophytes for negative MEI v.2 index anomalies, which represent La Niña conditions. For cyanobacteria and haptophytes, we observed a pronounced threshold response to positive anomalies, demonstrating sensitivity to ENSO events. Diatoms do not exhibit a clear threshold response for the MEI v.2 index but are affected across the entire range. Interestingly, positive anomalies in the MEI v.2, representing strong ENSO events, are mostly found in regime 1 (see Figure \ref{fig:clustering} $e$).

@@

These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separates the two regimes is NMDS 1, which covaries most strongly with changes in AMO, MEI v.2, salinity, evaporation, and precipitation (Figure \ref{fig:clustering} $c$). Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS 2. Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, show greater variance along NMDS 1. Salinity, evaporation, and silicate concentration exhibit some explanatory power in the gradient forest analysis, although they are weaker predictors than SST, nitrate concentration, and climate indices (see Figure \ref{fig:GF}). This implies that the climate indices explain a variability that more direct measurements of upwelling intensity do not capture.

@@

% 1.  Discuss Limitations of Light microscopy cell counts in CARIACO Time Series
Using microscopy cell counts as a basis for our analyses means that the results should be interpreted in terms of changes in the microphytoplankton community. Our choice of grouping cell counts by genus is based on the assumption that species within a genus generally exhibit similar eco-physiological responses. The output of the gradient forest model is further grouped by broad functional types to facilitate the identification of shifts and trends. The phytoplankton genera that can be best predicted are mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (Figure \ref{fig:GF} $b$). This does not represent a complete picture of the phytoplankton community of the CARIACO station. The HPLC pigment data used by \citeauthor{pinckney_phytoplankton_2015} capture all size classes of phytoplankton, but as previously discussed, does not cover the entire time series and the dataset was assembled by four different laboratories. With the caveat that the microscopy counts capture microphytoplankton exclusively, the data can be brought into agreement. Drastic declines in cell counts reflect shifts to smaller species, an aspect consistent with size spectra calculated by previous studies using HPLC pigment data \cite{lorenzoni_characterization_2015}.

% 2. Discuss limitations of GF analysis & Limitation to bottom-up drivers and climate indices
The restricted scope of the microscopy genus counts poses limitations to the gradient forest algorithm we employed to quantify predictors of community change.
Many of the predictors in ecological time series and in this study are highly correlated (Figure \ref{fig:sup:correlation} in Appendix). This can lead to the identification of a correlated predictor that has no direct effect on community changes. However, the gradient forest methodology addresses this by conditionally permuting subsets of the correlated predictors, which can partly mitigate this effect \cite{ellis_gradient_2012}. Additionally, random forest models do not consider temporal relationships between data points, which is why we included tests for the predictive relationships with time lags of our predictors. Without adjusting for time lags, SST and nitrate concentration are the strongest predictors of community change (Figure \ref{fig:sup:GFoutput_nolags} in Appendix). Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both SST and AMO index as the strongest predictors yield a relatively low value around \qty{0.03}{}. This is expected given the relatively noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the gradient forest algorithm \cite{pitcher_example_2012, roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}. It is important to consider that although predictive relationships between drivers and ecological predictors are identified, this does not necessarily imply a mechanistic relationship. The scope of our analysis, like any statistical analysis, is exploratory and aims at narrowing down potential key relationships. Mathematical modeling and experimental work are required to develop a clearer mechanistic understanding of the drivers of ecosystem change in the Cariaco Basin.

Our focus on bottom-up drivers of community change may also restrict the explanatory power of our analysis, as top-down processes such as zooplankton grazing may impact on the structure of the phytoplankton community \cite{banas_adding_2011, acevedo-trejos_mechanisms_2015}. Due to the scarcity of data covering higher trophic levels and the absence of zooplankton data for the first six years of the time series, we suggest that mechanistic modeling is the ideal approach for investigating the relative impact of top-down and bottom-up drivers.

\section{Conclusions}
We found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco Basin. Specifically, microscopy cell counts reveal the presence of two distinct regimes (and a return, during the final years, to regime 1) defined by high- and low-upwelling conditions that correspond to anomalies in the AMO index. We observe no significant changes in diversity metrics between the two regimes, except for lower turnover and higher community evenness under low-upwelling conditions.
Despite multiannual trends in warming, reduced productivity, and drastic declines in microphytoplankton cell counts, the phytoplankton community returned to regime 1 levels once the climatic and local physical conditions were restored. 
The strongest bottom-up variable predicting changes in the phytoplankton community resulted the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local conditions through low-frequency natural cycles. We concur with \citeA{nye_ecosystem_2014} that the AMO appears to be a suitable candidate as an indicator of ecosystem state for the phytoplankton community of the Cariaco Basin. 
Notwithstanding a number of limitations, the incipient return of the Cariaco basin to a pre-2004 regime hints at the resilient nature of phytoplankton communities under global change, an aspect that points to the relevance of long-term time series and deserves further attention. 
Future studies should focus on the relative impact of top-down and bottom-up drivers.

The analysis could be extended by including top-down effects on the phytoplankton community, particularly zooplankton and fisheries data. 
We hope that the work hints at the resilient nature of phytoplankton communities under global change, which should be further investigated by studying long-term ocean time series.

>>>>

\begin{keypoints}
\item The CARIACO phytoplankton community time series reveals two clusters, separated by a shift from high- to low-upwelling conditions after 2004 and a return to high-upwelling conditions following 2013.
\item No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low-upwelling conditions.
\item Gradient forest analysis identifies the AMO climate index as the strongest predictor of community change, highlighting the influence of large-scale climatic oscillations on this tropical coastal ecosystem.
\end{keypoints}

@@

\begin{abstract}
Phytoplankton communities play a crucial role in oceans' biogeochemical cycles. Climate drives changes in the physical environment that phytoplankton communities experience, thus impacting the productivity and diversity of marine ecosystems. Long-term time series can capture changes in climate and ecosystems, allowing us to untangle the effects on local communities. 
This study aims to describe and investigate the changes in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017. Water samples were taken during monthly cruises to determine microscopy cell counts and biogeochemical properties, which we integrated with data from the AMO and MEI v.2 climate indices and ERA5 climate model reanalysis data. 
Using clustering algorithms, we investigate the structure of the community data, revealing two clusters that describe a shift from high- to low-upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014.
However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009. To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related variables (SST and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors. Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of ecosystems to recover from the effects of climate change.

@@

    A tropical coastal location that has been studied intensively in the last few decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters have been consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}. 
    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world oceans \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression comprising two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated. This underwater topography, in combination with high surface productivity and aerobic degradation of sinking organic matter, creates permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for understanding the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.

@@

    The monthly AMO index data (10-year low-pass), which represents area-averaged, detrended low-pass filtered North Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.

@@

\subsection{Data analysis}    
    To provide an overview of the available data and the general trends, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$). Cruises from 1995, 2016, and 2017 were excluded because insufficient coverage in these years prevented us from producing reliable yearly mean estimates.

@@

To check whether the annual aggregation of community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (Figure \ref{fig:clustering}$c$). The NMDS clustering is overlaid with environmental data and climate indices in an attempt to unveil some of the underlying trends. The horizontal axis separates the two clusters, with cluster 2 located over positive anomalies in AMO and precipitation, and cluster 1 located over positive anomalies in evaporation, salinity, and MEI v.2. The second NMDS axis appears to follow the trends in upwelling intensity, as indicated by the vectors pointing in opposite directions for nutrient concentrations, temperature, and wind speed. Overall, the NMDS clustering confirms the pattern shown by clustering on annually aggregated data.

@@

\subsection{Community reorganization and biomass shifts}
% P1: Introductory paragraph highlighting the overall dynamics/shifts
This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations impacted the wind-driven upwelling regime, producing confounding effects on the composition and biodiversity of the phytoplankton community.
These changes are well visible in the yearly mean anomalies (Figure \ref{fig:zscore}) and the density distribution of monthly measurements (Figure \ref{fig:clustering}). After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies up to 2014. 
By considering the phytoplankton community time series until the end of 2017, the temporally complete dataset, we find that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}. In 2014, physical conditions in the Cariaco Basin returned to a regime of increased wind speed, increased upwelling, and elevated nutrient concentrations (Figure \ref{fig:zscore}). Concurrently, we observe an increase in chlorophyll \textit{a} (Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (Figure \ref{fig:clustering} $a$ and $c$).

% P2: Discuss community reorganisation, biomass shifts vs HPLC data 
When comparing the two regimes, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a}, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}). The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community might have transitioned to smaller cells, which were probably below the detection size limit of optical microscopes \cite{muller-karger_scientific_2019}. This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}. Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but shifts towards smaller cells are particularly prevalent when temperatures increase and nutrient supply decreases \cite{mousing_global_2014}, as we observed in the Cariaco Basin.

% Disucssion of HPLC vs Microscopy!
A similar pattern emerges when comparing our observations with High-Performance Liquid Chromatography (HPLC) data from the Cariaco Basin, which were analyzed and presented by \citeA{pinckney_phytoplankton_2015}. The HPLC data suggest an increase in chlorophyll \textit{a} between the periods 1996-2000 and 2006-2010, particularly at depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}. Despite cell counts of all functional groups decreasing, the HPLC data indicate an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{\meter}. This is a further strong indication that the phytoplankton community shifted to smaller sizes, undetected using light microscopy.
It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006. \citeA{pinckney_phytoplankton_2015} could not find a consistent method for quantifying total chlorophyll \textit{a} across different laboratories. This is why we focused our analysis on the more consistent and temporally complete microscopy cell count and fluorometric chlorophyll \textit{a} data.
Consistent with our results, independent spectrophotometric measurements of chlorophyll \textit{a} concentrations at \qty{20}{meters} depth, off the coast of Margarita Island, show a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.

% P3: Shift in Diversity does not match shift in biomass/cell counts, why?
The temporal dynamics we find in yearly mean genus richness, however, do not correspond directly to the observed regimes. Instead, we see a marked reduction from 1998 to 2010, after which richness recovers four years prior to 2014, when the community returns to levels characteristic of regime 1. The same trend is evident, albeit to a lesser degree, in the other diversity metrics. Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and exhibit a reduced yearly mean between 2000 and 2006, during which cell counts recorded an increase in nanoflagellates and a pronounced peak in cyanobacteria.
We did not find a significant difference in diversity between the two regimes, either by Shannon index or by genus richness (Figures \ref{fig:zscore} and \ref{fig:divts}). However, we did observe a significant difference in Pielou's index between the two regimes. The results suggest that the phytoplankton community becomes more evenly distributed during low-upwelling conditions of regime 2, which is consistent with the analysis of \citeauthor{pinckney_phytoplankton_2015} based on HPLC pigment data. This increase in evenness, quantified by the Pielou index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community. The resulting decrease in compositional turnover is also observed in a decline in interannual species turnover during regime 2 (see Figure \ref{fig:clustering}$b$). As the productivity and biomass of phytoplankton communities are driven primarily by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.

@@

% P5: Climate impacts
Although regional effects may vary, a negative phase in the AMO is generally associated with periods of colder temperatures and a southward shift of the ITCZ \cite{knight_climate_2006, colna2017latitudinal}. In the Cariaco Basin, a southward shift of the ITCZ correlates with increased wind-driven upwelling \cite{taylor_ecosystem_2012}. The threshold responses between functional groups, which relate to shifts in cell count abundances, show contrasting patterns. Diatoms, dinoflagellates, and haptophytes exhibit strong response thresholds for negative AMO anomalies, which could point to the influence of stronger upwelling conditions. The abundance of haptophytes, additionally, shows a threshold effect at positive AMO anomalies, which may be related to the pattern observed by \citeauthor{pinckney_phytoplankton_2015} based on HPLC data that haptophytes predominantly exhibit blooms above the mixed layer and at a lower frequency than other functional groups.

The El Niño Southern Oscillation (ENSO) exhibits a strong teleconnection to the Caribbean Sea. During ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months. In contrast, we found a relatively strong predictive power for changes in the phytoplankton community for the MEI v.2 index with a 4-month time lag. We also tested a 12-month lag and found no significant improvement (Table \ref{tab:sup:GFlagtests} in Appendix).
The threshold responses indicate a shift in the abundances of dinoflagellates and haptophytes for negative MEI v.2 index anomalies, which represent La Niña conditions. For cyanobacteria and haptophytes, we observed a pronounced threshold response to positive anomalies, demonstrating sensitivity to ENSO events. Diatoms do not exhibit a clear threshold response for the MEI v.2 index but are affected across the entire range. Interestingly, positive anomalies in the MEI v.2, representing strong ENSO events, are mostly found in regime 1 (see Figure \ref{fig:clustering} $e$).

@@

These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separates the two regimes is NMDS 1, which covaries most strongly with changes in AMO, MEI v.2, salinity, evaporation, and precipitation (Figure \ref{fig:clustering} $c$). Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS 2. Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, show greater variance along NMDS 1. Salinity, evaporation, and silicate concentration exhibit some explanatory power in the gradient forest analysis, although they are weaker predictors than SST, nitrate concentration, and climate indices (see Figure \ref{fig:GF}). This implies that the climate indices explain variability that more direct measurements of upwelling intensity do not capture.

@@

% 1.  Discuss Limitations of Light microscopy cell counts in CARIACO Time Series
Using microscopy cell counts as a basis for our analyses means that the results should be interpreted in terms of changes in the microphytoplankton community. Our choice of grouping cell counts by genus is based on the assumption that species within a genus generally exhibit similar eco-physiological responses. The output of the gradient forest model is further grouped by broad functional types to facilitate the identification of shifts and trends. The phytoplankton genera that can be best predicted are mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (Figure \ref{fig:GF} $b$). This does not represent a complete picture of the phytoplankton community of the CARIACO station. The HPLC pigment data used by \citeauthor{pinckney_phytoplankton_2015} capture all size classes of phytoplankton, but, as previously discussed, it does not cover the entire time series and the dataset was assembled by four different laboratories. With the caveat that the microscopy counts capture microphytoplankton almost exclusively, the discrepancies to the fluorometric chlorophyll \textit{a} measurements can be explained. Drastic declines in cell counts reflect shifts to smaller species, an aspect consistent with size spectra calculated by previous studies using HPLC pigment data \cite{lorenzoni_characterization_2015}.

% 2. Discuss limitations of GF analysis & Limitation to bottom-up drivers and climate indices
The limited scope of the microscopy cell counts also affects the conclusions that can be drawn from the Gradient Forest analysis, as these results can only be interpreted as concerning microphytoplankton. 
Another potential caveat of our methodology is that many of the predictors in ecological time series and in this study are highly correlated (Figure \ref{fig:sup:correlation} in Appendix). This can lead to the identification of a correlated predictor that has no direct effect on community changes. However, the gradient forest methodology addresses this by conditionally permuting subsets of the correlated predictors, which can partly mitigate this effect \cite{ellis_gradient_2012}. Additionally, random forest models do not consider temporal relationships between data points, which is why we included tests for the predictive relationships with time lags of our predictors. Without adjusting for time lags, SST and nitrate concentration are the strongest predictors of community change (Figure \ref{fig:sup:GFoutput_nolags} in Appendix). Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both SST and AMO index as the strongest predictors yield a relatively low value around \qty{0.03}{}. This is expected given the relatively noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the gradient forest algorithm \cite{pitcher_example_2012, roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}. It is important to consider that although predictive relationships between drivers and ecological predictors are identified, this does not necessarily imply a mechanistic relationship. The scope of our analysis, like any statistical analysis, is exploratory and aims at narrowing down potential key relationships. Mathematical modeling and experimental work are required to develop a clearer mechanistic understanding of the drivers of ecosystem change in the Cariaco Basin.

Our focus on bottom-up drivers of community change may also restrict the explanatory power of our analysis, as top-down processes such as zooplankton grazing may impact the structure of the phytoplankton community \cite{frank_ups_2007, banas_adding_2011, acevedo-trejos_mechanisms_2015}. Due to the scarcity of data covering higher trophic levels and the absence of zooplankton data for the first six years of the time series, we suggest that mechanistic modeling is the ideal approach for investigating the relative impacts of both top-down and bottom-up drivers.

\section{Conclusions}
We found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco Basin. Specifically, microscopy cell counts reveal the presence of two distinct regimes defined by high- and low-upwelling conditions that correspond to anomalies in the AMO index. The community changes do not indicate a permanent regime shift, as a return to regime 1 is observed towards the end of the time series. We observe no significant changes in diversity metrics between the two regimes, except for lower turnover and higher community evenness under low-upwelling conditions.
Despite multiannual trends in warming, reduced productivity, and drastic declines in microphytoplankton cell counts, the phytoplankton community returned to previous levels once the climatic and local physical conditions were restored. 
The strongest bottom-up variable predicting changes in the phytoplankton community was the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local conditions through low-frequency natural cycles. We concur with \citeA{nye_ecosystem_2014} that the AMO appears to be a suitable candidate as an indicator of ecosystem state for the phytoplankton community of the Cariaco Basin. 
Notwithstanding a number of limitations, our analysis points to the incipient return of the Cariaco Basin to a pre-2004 regime, hinting at the resilient nature of phytoplankton communities under global change, an aspect that emphasizes the relevance of long-term time series and deserves further attention. 
To extend our approach, future studies should include the impact of top-down drivers and investigate their explanatory power regarding the phytoplankton community changes observed in the Cariaco Basin.
